Insecticide resistance evolution with mixtures and sequences : a model-based explanation.

### 2018-01-02 inst/documents/paper2/paper2_resistance_mechanisms_mixtures.Rmd
new section on starting frequencies

#### Starting frequencies of resistance
Changing the starting frequency of resistance had a similar effect on time to resistance for both sequences and mixtures and thus had little effect on their relative performance. For example taking the base scenario and reducing the starting frequency of one resistance allele did not change from sequence being favoured (compare Fig 9C to 9A). Similarly taking a scenario in which time-to-resistance is longer for a mixture (Fig 9B) and decreasing the starting frequency of that resistance allele did not change the fact that the mixture was favoured (Fig 9D).

@@

#### Linkage disequilibrium
Positive linkage disequilibrium is said to exist when the frequency of two alleles at different loci is higher than expected if they were just selected individually. It has previously been suggested [@Curtis1985][@WHO2012] that using an insecticide to which there was existing resistance in a mixture with a new insecticide with little resistance could lead to more rapid evolution of resistance to the latter through linkage disequilibrium. We explicitly look at the example of a mixture of a new insecticide and one with pre-existing resistance (Figures 9 C&D). In both cases resistance to the newer insecticide with lower starting resistance (in red) increases faster when it is used alone (red dotted) than when it is used in a mixture (red solid). This suggests that linkage disequilibrium does not restrict the potential benefits of mixtures in all situations. These results support the statement in GPIRM [1] that : "Even though the risk for linkage disequilibrium exists, mixtures may still be the most attractive IRM tool …". The dynamics of LD is too complex to discuss in detail here but interested readers can find more discussion in [@Curtis1985][@WHO2012][@Levick2017].

@@

There are other model behaviours that can be changed by modifying model inputs within the computer code. For example, here we assume that insecticide interaction in a mixture is multiplicative, e.g. if the probability of surviving exposure to insecticide A alone is 0.3 and of surviving insecticide B alone is 0.2, then the probability of surviving exposure to a mixture of A and B is 0.3*0.2 = 0.06. This is consistent with recent work on aphids [@Taillebois2016], although that did also indicate some synergistic effects. Modifying model inputs would allow us to incorporate such synergistic or antagonistic effects of mixtures, i.e. that insecticides used together perform better or worse than when used separately. This would allow the model to represent cross-resistance where a resistance gene provides resistance to more than one insecticide. Similarly behavioural resistance is likely to offer resistance against more than one insecticide [@GPIRM2012] but until there are more data on whether behavioural resistance is genetic and heritable we would not recommend representing it in this way.

>>>>

#### Starting frequencies of resistance
Changing the starting frequency of resistance had a similar effect on time to resistance for both sequences and mixtures and thus had little effect on their relative performance. For example taking the base scenario and reducing the starting frequency of one resistance allele did not change from sequence being favoured (compare Fig 9C to 9A). Similarly taking a scenario in which time-to-resistance is longer for a mixture (Fig 9B) and decreasing the starting frequency of that resistance allele did not change the fact that the mixture was favoured (Fig 9D). In these analyses we set starting frequencies of resistance relatively high at 0.01 or 0.001. In our more detailed sensitivity analysis [@Levick2017] the lower limit for starting frequencies was 0.0001 and still it was not a key parameter in determining the relative performance of mixtures and sequences.

@@

#### Linkage disequilibrium
Positive linkage disequilibrium is said to exist when the frequency of two alleles at different loci is higher than expected if they were just selected individually. It has previously been suggested [@Curtis1985][@WHO2012] that using an insecticide to which there was existing resistance in a mixture with a new insecticide with little resistance could lead to more rapid evolution of resistance to the latter through linkage disequilibrium. We explicitly look at the example of a mixture of a new insecticide and one with pre-existing resistance (Figures 9 C&D). In both cases resistance to the newer insecticide with lower starting resistance (in red) increases faster when it is used alone (red dotted) than when it is used in a mixture (red solid). This suggests that linkage disequilibrium does not restrict the potential benefits of mixtures in all situations. These results support the statement in GPIRM [@WHO2012] that : "Even though the risk for linkage disequilibrium exists, mixtures may still be the most attractive IRM tool …". The dynamics of LD is too complex to discuss in detail here but interested readers can find more discussion in [@Curtis1985][@WHO2012][@Levick2017].

@@

There are other model behaviours that can be changed by modifying model inputs within the computer code. For example, here we assume that insecticide interaction in a mixture is multiplicative, e.g. if the probability of surviving exposure to insecticide A alone is 0.3 and of surviving insecticide B alone is 0.2, then the probability of surviving exposure to a mixture of A and B is 0.3*0.2 = 0.06. This is consistent with recent work on aphids [@Taillebois2016], although that did also indicate some synergistic effects. Modifying model inputs would allow us to incorporate such synergistic or antagonistic effects of mixtures, i.e. that insecticides used together perform better or worse than when used separately. This would allow the model to represent cross-resistance where a resistance gene provides resistance to more than one insecticide. Similarly behavioural resistance is likely to offer resistance against more than one insecticide [@WHO2012] but until there are more data on whether behavioural resistance is genetic and heritable we would not recommend representing it in this way.
